---
title: Perceptual awareness for somatosensory stimuli varies across the cardiac cycle
author: "Pawel Motyka"
contact: "pawel.motyka@psych.uw.edu.pl"
date: "August 25th 2017"
output: html_document
---

List of sections:

1. Load required packages, behavioral data, and perform exclusions [S1](#S1)
2. Classify behavioral responses [S2](#S2)
3. Calculate the relative position of stimulus onset within the RR intervals [S3](#S3)
4. Test unifromity of stimulus onsets distribution within the RR intervals for each participant [S4](#S4)
5. Analyze behavioral data [S5](#S5)
6. Test unifromity of hits distribution within the RR intervals (Figure 2A) [S6](#S6)
7. Test unifromity of misses distribution within the RR intervals (Figure 2B) [S7](#S7)
8. Calculate sensitivity (d') and response criterion (c) for systole and diastole [S8](#S8)
9. Test differences in sensitivity between systole and diastole (Figure 3A) [S9](#S9)
10. Test differences in response criterion between systole and diastole (Figure 3B) [S10](#S10)

--- 
<a name="S1"></a>
&nbsp;

**1. Loading required packages, behavioral data, and perform exclusions**


```{r, results = "hide", message = FALSE,  warning = FALSE}

# load required packages
require(circular) 
require(R.matlab) 
require(Scale) 
require(ggplot2)
require(scales)

# set up data directory
data_dir <- 'N:/CCSomato_data'
setwd(data_dir)

# load behavioral data
data <- read.table("CCSomato_trials.txt", header = TRUE, sep = "\t", fill = TRUE, stringsAsFactors = FALSE)

## list of the crucial variables in behavioral data file:
  # "stim_type" - stimulus type [1 - near-treshold, 0 - catch trial]
  # "resp1" - response [1 - "yes", 0 - "no"] 
  # "resp1_t" - response time [in seconds]
  # "stim_onset" - stimulus onset time [in seconds]

# exclude participant 01 (one of the authors tested during final pilot trial)
data <- data[!(data$ID == 1),] 

# exclude one block due to an ECG recording error in this block
data <- data[!(data$ID == 4 & data$block == 1), ] 

```
<a name="S2"></a>
&nbsp;

**2. Classify behavioral responses**

```{r, results = "hide", message = FALSE}

# classify as a hit in case of the stimulus present and "yes" response
data$resp_class[data$stim_type == 1 & data$resp1 == 1] <- 'hit' 

# classify as a miss in case of the stimulus present and "no" response
data$resp_class[data$stim_type == 1 & data$resp1 == 0] <- 'miss' 

# classify as a false alarm in case of the catch trial and "yes" response
data$resp_class[data$stim_type == 0 & data$resp1 == 1] <- 'FA' 

# classify as a correct rejection in case of the catch trial and "no" response
data$resp_class[data$stim_type == 0 & data$resp1 == 0] <- 'CR' 

# save response class variable as a factor
data$resp_class <- factor(data$resp_class)

```
<a name="S3"></a>
&nbsp;

**3. Calculate the relative position of stimulus onsets within RR intervals**

```{r}

# specify ECG files prefix
file_prefix <- 'CCSomato_'

# add variable "ao_error" specfiying whether recording error has occurred [1 - "yes", 0 - "no"] - while collecting data from 12th participant, the data acquisition card produced an error (and stopped the recording). Since then, we recorded the occurence of the error. The error occurred later in case of 15 trials that have been later excluded. For the first 11 participants, when the card did not crashed, we assigned value 0.
data$ao_error[is.na(data$ao_error)] <- 0 

# specify a dataframe with the number of ECG events

ecg_events_list <- data.frame(ID = integer(0),
                              block = integer(0),
                              ecg_event_num = numeric(0))

### LOOP PARTICIPANTS (p) 

for (p in unique(data$ID)) {ifelse(p < 10, ID_str <- paste('0', p, sep = ''), ID_str <- p)  
    
  ## LOOP BLOCKS (b)
  
    for (b in unique(data$block[data$ID==p])) {
    
    # specify directories containing ECG data
    kubios_mat <- paste(data_dir, '/ID', ID_str, '/ecg/kubios/', file_prefix, ID_str, '_exp_0', b,'_hrv','.mat', sep='')
    ecg_event_file <- paste(data_dir, '/ID', ID_str, '/ecg/eeg_lab/', file_prefix, ID_str, '_events_exp_0', b, '.csv', sep='')
    
    # specify ECG sampling frequency (in Hz)
    ecg_sf <- 1000
    
    # load the vectors with R-peaks time series from Kubios matlab files into R
    R_peaks_tmp <- readMat(kubios_mat)
    R_peaks <- as.vector(R_peaks_tmp$Res[[4]][[2]][[2]]) # the numbers in the list reflect the position of R peaks time variable in Kubios matlab files
    rm(R_peaks_tmp)
    
    # import markers of events/responses from ECG file
    ecg_events <- read.csv(ecg_event_file, sep = '\t')
    
    # import a vector with stimulus onset time series (in seconds)
    stim_onsets <- ecg_events$latency[ecg_events$type=='S  1']/ecg_sf
    
    # subset a vector with the sequence of trials (stimulation or catch trial)
    sequence <- subset(data, ID==p & block==b, stim_type)
    
    # subset a vector with the sequence of responses ("yes" or "no")
    responses <- subset(data, ID==p & block==b, resp1)
    
        # LOOP TRIALS (tr) 
    
        # introduce variables computed in the loop on the trial level (described below within the loop)
        pos <- c()
        diff2peak <- c()
        diff2peak_deg <-c()
        diff2peak_abs <-c()
        RR_interval <- c()
        stim_degree <- c()
        ecg_i <- 0
    
        for (tr in 1:120) { # 120 - number of trials per block
          
          # perform further computations while skipping the trials when the ao_error occurred
          if (data$ao_error[data$ID == p & data$block == b & data$trial == tr] == 1) 
          {data$diff2peak_deg[data$ID == p & data$block == b & data$trial == tr] <- NA} 
          else {ecg_i <- ecg_i+1
            
              # encode the time of the R peak before the stimulus onset
              pos[ecg_i] <- max(which(R_peaks < stim_onsets[ecg_i]))
              
              # encode the difference between the stimulus onset time and the previous R peak
              diff2peak[ecg_i] <- stim_onsets[ecg_i] - R_peaks[pos[ecg_i]]
              
              # encode tha duration of the RR interval during which the stimulus occurred
              RR_interval[ecg_i] <- (R_peaks[pos[ecg_i]+1] - R_peaks[pos[ecg_i]])
              
              # encode the relative position of stimulus onset within the RR interval, assigning the values from 0 to 360 degrees (with 0 indicating the R peak previous to the stimulus onset). 
              stim_degree[ecg_i] <- 360 * diff2peak[ecg_i]/(R_peaks[pos[ecg_i]+1] - R_peaks[pos[ecg_i]])
              
              # add a variable specifying the relative position of stimulus within the RR interval
              data$diff2peak_deg[data$ID == p & data$block == b & data$trial == tr] <- stim_degree[ecg_i]
              
              # add a variable specifying the absolute difference between the stimulus onset and the previous R peak (in seconds)
              data$diff2peak_abs[data$ID == p & data$block == b & data$trial == tr] <- diff2peak[ecg_i]
              
              # add the "RR interval" variable (in seconds)
              data$RR_interval[data$ID == p & data$block == b & data$trial == tr] <- RR_interval[ecg_i] 
      
              } # END: perform computations skipping the trials when the ao_error occurred
      
        } # END: LOOP TRIAL (tr)
    
    # Create data frame with number of valid ECG events - introduced for controlling purposes
    ecg_events_list[nrow(ecg_events_list)+1,] <- c(p,b,length(stim_onsets))
    
    } ## END: LOOP BLOCKS (b)
  
} ### END: LOOP PARTICIPANTS (p)

# remove the objects used within the loop
rm(responses)
rm(ecg_events)
rm(sequence)

# remove 15 trials in which ao_error occured and onset position was not calculated
data <- data[!(data$ao_error == 1),]

# remove 2 trials with outlying RR intervals lengths (i.e. ~2.1 and ~3.4 seconds) observed due to missed fragments of ECG signal. The third longest interval equaled ~ 1.26 sec.
length(data$RR_interval[data$RR_interval > 1.27])
data <- data[!(data$RR_interval > 1.27),]

# remove 191 trials in which participants did not give any response within 2 seconds 
length(data$resp1_t[data$resp1_t >= 2])
data <- data[!(data$resp1_t >= 2),]

# show remaining number of the trials after exclusions
length(data$trial)

## Saving and loading the processed data
#write.table(data, file = "CCSomato_processed_data(output_from_section_3)", sep = "\t")
#data <- read.table("CCSomato_processed_data(output_from_section_3)", header = TRUE, sep = "\t", fill = TRUE, stringsAsFactors = FALSE)

```
<a name="S4"></a>
&nbsp;

**4. Test unifromity of stimulus onsets distribution within the RR intervals for each participant**
<a name="S4"></a>

```{r, fig.width=4, fig.height=4}

## General information about the Rayleigh Test (Circular Statistics)
#"Rayleigh Test of Uniformity" assess the significance of the mean resultant length (—R) which is "a measure of the concentration of data points around the circle. If —R is greater than a threshold value, the data are interpreted as being too concentrated to be consistent with uniformity" (Pawsey et al., 2013, s. 81). The alternative hypothesis is a unimodal distribution with unknown mean direction. For more details on using circular statistics in R please see: https://cran.r-project.org/web/packages/circular/circular.pdf

# specify a dataframe with the uniformity test results for each participant
Onset_distribution_ID <- data.frame(ID = integer(0),
                              test_statistics = numeric(0),
                              p_value = numeric(0))

for (p in unique(data$ID)) { # LOOP PARTICIPANTS (p)
  
  # save the vector with calculated onset positions in degrees as a circular object
  temp_onset <- circular(data$diff2peak_deg[data$ID==p], type="angles", units="degree", rotation="clock", zero=pi/2) 

  
  # plot the distribution of stimulus onsets within the RR intervals for each participant - with 0 indicating the R peak before stimulus onset
  plot(temp_onset, stack=TRUE, col = "gray25", cex = 0.77, main = paste("Onset distibution \n participant nr", p))
  rose.diag(temp_onset, bins = 16, col = "cadetblue4", prop = 2.5, add = T, axes = F)
  circ.dens = density(temp_onset, bw=40)
  lines(circ.dens, col="lightcyan3", lwd = 3, xpd=TRUE)
  
  # test uniformity of onset distribution for a particpant and save the results
  onset_distribution_individual <- rayleigh.test(temp_onset) 

  # show test statistics and p-value for a participant
  participant <- paste("ID", p)
  print(participant)
  print(onset_distribution_individual)  
  
  # add test statistics and p-value to the dataframe with uniformity testing results
  p_value <- onset_distribution_individual$p.value
  test_statistics <- onset_distribution_individual$statistic
  Onset_distribution_ID[nrow(Onset_distribution_ID)+1,] <- c(p, test_statistics, p_value)
  
}

# Rayleigh Test yields significant result in case of participant nr 12 indicating a non-uniform distribution of stimulus onsets across the cardiac cycle (RR intervals)

# remove participant nr 12
data <- data[data$ID != 12,] 

# show a total number of valid trials
length(data$trial)
length(data$trial[data$resp_class == "hit"])
length(data$trial[data$resp_class == "miss"])
length(data$trial[data$resp_class == "FA"])
length(data$trial[data$resp_class == "CR"])

```
<a name="S5"></a>
&nbsp;

**5. Analyze behavioral data**

```{r}

# Specify a dataframe with the set of behavioral variables
data_behav <- data.frame(ID = integer(0),
              age = integer(0),
              hit = numeric(0),
              miss = numeric(0),
              FA = numeric(0), 
              CR = numeric(0),  
              hit_rate = numeric (0),
              FA_rate = numeric(0)) 

# LOOP PARTICIPANTS (p) 

for (p in unique(data$ID)) {
  
     #age
     age <- unique(data$age[data$ID == p])
     
     # count hits
     hit <- length(data$resp_class[data$resp_class =="hit" & data$ID == p])
     
     # count misses
     miss <- length(data$resp_class[data$resp_class =="miss" & data$ID == p])
     
     # count false alarms
     FA <- length(data$resp_class[data$resp_class =="FA" & data$ID == p ])
     
     # count correct rejections
     CR <- length(data$resp_class[data$resp_class =="CR" & data$ID == p ])
     
     # calculate hit rate (varying from 0 to 1)
     hit_rate <- (length(data$resp_class[data$resp_class =="hit" & data$ID == p]))/(length(data$trial[data$ID == p & data$stim_type == 1]))
     
     # calculate false alarms rate (varying from 0 to 1)
      FA_rate <- (length(data$resp_class[data$resp_class =="FA" & data$ID == p]))/(length(data$trial[data$ID == p & data$stim_type == 0]))
      
     # create a dataframe with the abovementioned variables
      data_behav[nrow(data_behav)+1,] <- c(p, age, hit, miss, FA, CR, hit_rate, FA_rate)
      
    } # END: LOOP PARTICIPANTS (p)

# show basic statistics for hit rate    
mean(data_behav$hit_rate)
median(data_behav$hit_rate)
sd(data_behav$hit_rate)

# show basic statistics for false alarms rate
mean(data_behav$FA_rate)
median(data_behav$FA_rate)
sd(data_behav$FA_rate)

# show basic statistics for age
mean(data_behav$age)
range(data_behav$age)
sd(data_behav$age)

```
<a name="S6"></a>
&nbsp;

**6. Test uniformity of hits distribution within the RR intervals**

```{r}

# specify a variable with the relative positions of hits within the RR intervals
mean_degree_hit <- vector()

# extract the list of participants
ID_list <- unique(data$ID)

for (p in ID_list) { # LOOP PARTICIPANTS (p)
  
  # save the vector with hits positions in degrees as a circular object
  temp_hit <- circular(data$diff2peak_deg[data$ID==p & data$resp_class == "hit"], type="angles", units="degree", rotation="clock", zero=0)
  
  # calculate the mean phase when hits occurred for each participant
  mean_degree_hit[p] <- mean(temp_hit)
  
} # END: LOOP PARTICIPANTS (p)

# save the vector with mean hits for each particiapant as a circular object
Hits_secondlevel <- circular(mean_degree_hit[ID_list], type="angles", units="degree", rotation="clock", zero=pi/2)

# plot the distribution of mean hits within the RR intervals - with 0 indicating the R peak previous to the stimulus onset
plot(Hits_secondlevel, stack= TRUE, bins = 720, col = "gray25", cex = 0.77, lwd = 2, main = "Hits")

# optionally add 'type = "n"' argument to eliminate points from the plot

# add lines representing mean hits for each participant
arrows.circular(Hits_secondlevel, col = rgb(0.23,0.48,0.50, 0.7), lwd = 2, length = 0.001)

# add an arrow representing grand mean for hits - with its length signifying resultant length of the grand mean
arrows.circular(mean(Hits_secondlevel), y=rho.circular(Hits_secondlevel), lwd = 3, col = rgb(0.13,0.28,0.30), length = 0.1)

# add line signifying distribution of hits (bandwidth = 20)
circ.dens = density(Hits_secondlevel, bw=20)
lines(circ.dens, col= rgb(0.38,0.63,0.65), lwd = 2, xpd=TRUE)

# test uniformity of hits distribution within the RR intervals
rayleigh.test(Hits_secondlevel)

# show the angle of grand mean for hits
mean(Hits_secondlevel)

```
<a name="S7"></a>
&nbsp;

**7. Test uniformity of misses distribution within the RR intervals**

```{r}

# specify a variable with the relative positions of misses within the RR intervals
mean_degree_miss <- vector()
ID_list <- unique(data$ID)

for (p in ID_list) { # LOOP PARTICIPANTS (p)

  # save the vector with the mean misses for each particiapant as a circular object
  temp_miss <- circular(data$diff2peak_deg[data$ID==p & data$resp_class == "miss"], type="angles", units="degree",  rotation="clock", zero=0)
  
  # calculate the mean phase when misses occurred for each participant
  mean_degree_miss[p] <- mean(temp_miss)

} # END: LOOP PARTICIPANTS (p)

# save the vector with mean misses for each particiapant as a circular object
Miss_secondlevel <- circular(mean_degree_miss[ID_list], type ="angles", units ="degree", rotation ="clock", zero=pi/2)

# plot the distribution of mean misses within the RR intervals - with 0 indicating the R peak previous to the stimulus onset
plot(Miss_secondlevel, stack=TRUE, bins = 720, col = "gray25", cex = 0.77, lwd = 2, main = "Misses")

# add lines representing mean misses for each participant
arrows.circular(Miss_secondlevel, col = rgb(0.23,0.48,0.50, 0.7), lwd = 2, length = 0.001)

# add an arrow representing grand mean for misses - with its length signifying resultant length of the grand mean
arrows.circular(mean(Miss_secondlevel), y=rho.circular(Miss_secondlevel), lwd = 3, col = rgb(0.13,0.28,0.30), length = 0.1)

# add line signifying distribution of misses (bandwidth = 20)
  circ.dens = density(Miss_secondlevel, bw=20)
  lines(circ.dens, col= rgb(0.38,0.63,0.65), lwd = 2, xpd=TRUE)
  
# test uniformity of misses distribution within RR intervals
rayleigh.test(Miss_secondlevel)

# show the angle of grand mean for misses
mean(Miss_secondlevel)

```
<a name="S8"></a>
&nbsp;

**8. Calculate sensitivity (d') and response criterion (c) for systole and diastole**


```{r}

## Binning of the stimulus onsets (in ms after the previous R-peak) into systole and diastole phase

# Classify the stimulus onsets delivered between R+100ms and R+400ms as delivered during systole phase
data$c_phase[data$diff2peak_abs >= 0.1 & data$diff2peak_abs <= 0.4] <- "systole"

# Classify the stimulus onsets delivered after R+400ms and between R+0 and R+100ms as delivered during diastole phase
data$c_phase[data$diff2peak_abs < 0.1 | data$diff2peak_abs > 0.4] <- "diastole"

# The applied binning ranges were informed by the procedures from other published studies;
  # Gray et al., (2012) systole: from R+100ms to R+500ms / diastole: from R-200ms to R+50ms 
  # Garfinkel et al. (2014) systole: from R+200ms to R+550ms / diastole: from R-300ms to R+175ms 
  # Yang et al. (2017) systole: R+180ms / diastole: R+480ms
  # Pramme et al.(2016) systole: R+185ms / diastole: R+480ms
  # Edwards et al.(2009) systole: R+300ms / diastole: R+0ms & R+600ms
  # Azevedo et al. (2017) systole: R+300ms / diastole: R+500ms
# Notably, the abovementioned intervals are not fully convergent between studies, so the present ranges reflect an approximation of the usal timing ascribed to systole and diastole

## Sensitivity and response criterion

# The sensitivity (d') and response criterion (criterion, c) parameters are calculated using a log-linear correction for extreme false alarm proportions (Hautus, 1995) since many participants produced no false alarms. Thus, the hit rate and false alarm rate are recalculated in the subsequent loop applying the correction: "H = (h + 0.5)/(Ns + 1) and F = (f + 0.5)/(NN + 1), where H and F refer to the frequency of hits and false alarms, respectively" (Hautus and Lee, 1998, s. 639; The dispersions of estimates of sensitivity obtained from four psychophysical procedures: Implications for experimental design; Perception & Psychophysics 1998, 60 (4), 638-649)

data_ID <- data.frame(ID = integer(0),
                 hit_rate = numeric (0),
                 FA_rate = numeric(0),
                 s = numeric(0), # sensitivity (d')
                 c = numeric (0), # criterion (c)
                 hit_rate_s = numeric(0), # hit rate at systole
                 hit_rate_d = numeric(0), # hit rate at diastole
                 FA_rate_s = numeric(0), # false alarm rate at systole
                 FA_rate_d = numeric(0), # false alarm rate at diastole
                 s_systole = numeric(0), # sensitivity (d') at systole
                 s_diastole = numeric(0), # sensitivity (d') at diastole
                 c_systole = numeric(0), # criterion (c) at systole
                 c_diastole = numeric (0)) # criterion (c) at diastole

for ( p in unique(data$ID)) { # LOOP PARTICIPANTS (p)
    
    # calculate hit rate
    hit_rate <- (length(data$resp_class[data$resp_class =="hit" & data$ID == p])+0.5)/(length(data$trial[data$ID == p & data$stim_type == 1])+1)
    
    # calculate false alarm rate
    FA_rate <- (length(data$resp_class[data$resp_class =="FA" & data$ID == p])+0.5)/(length(data$trial[data$ID == p & data$stim_type == 0])+1)
    
    # calculate sensitivity (d') [d' = Z(hit_rate) - Z(False_alarms_rate)] 
    s <- qnorm(hit_rate) - qnorm(FA_rate) 
    
    # calculate criterion (c) [c = - [Z(hit_rate) + Z(False_ alarms_rate)]/2] 
    c <- -(qnorm(hit_rate)+qnorm(FA_rate))/2 
    
    # calculate hit rate at systole
    hit_rate_s <- (length(data$resp_class[data$resp_class =="hit" & data$ID == p & data$c_phase == "systole"])+0.5)/(length(data$trial[data$ID == p & data$stim_type == 1 & data$c_phase == "systole"])+1)
    
    # calculate hit rate at diastole
    hit_rate_d <- (length(data$resp_class[data$resp_class =="hit" & data$ID == p & data$c_phase == "diastole"])+0.5)/(length(data$trial[data$ID == p & data$stim_type == 1 & data$c_phase == "diastole"])+1)
    
    # calculate false alarm rate at systole
    FA_rate_s <- (length(data$resp_class[data$resp_class =="FA" & data$ID == p & data$c_phase == "systole"])+0.5)/(length(data$trial[data$ID == p & data$stim_type == 0 & data$c_phase == "systole"])+1)
    
    # calculate false alarm at diastole
    FA_rate_d <- (length(data$resp_class[data$resp_class =="FA" & data$ID == p & data$c_phase == "diastole"])+0.5)/(length(data$trial[data$ID == p & data$stim_type == 0 & data$c_phase == "diastole"])+1)
    
    # calculate sensitivity (d') at systole
    s_systole <- qnorm(hit_rate_s) - qnorm(FA_rate_s) 
    
    # calculate sensitivity (d') at diastole
    s_diastole <- qnorm(hit_rate_d) - qnorm(FA_rate_d) 
    
    # calculate criterion (c) at systole
    c_systole <- -(qnorm(hit_rate_s)+qnorm(FA_rate_s))/2 
    
    # calculate criterion (c) at diastole
    c_diastole <- -(qnorm(hit_rate_d)+qnorm(FA_rate_d))/2 
    
    # create a dataframe with sensitivity and criterion measures
    data_ID[nrow(data_ID)+1,] <- c(p, hit_rate, FA_rate,s,c,hit_rate_s,hit_rate_d,FA_rate_s,FA_rate_d, s_systole,s_diastole,c_systole,c_diastole)
    
} # END: LOOP PARTICIPANTS (p)


```
<a name="S9"></a>
&nbsp;

**9. Test differences in sensitivity between systole and diastole**

```{r}

## Sensitivity (d') at systole and diastole

# test normality of sensitivity distribution at systole and diastole
shapiro.test(data_ID$s_systole)
shapiro.test(data_ID$s_diastole)

# employ paired t-test to determine whether sensitivity differs between the systole and diastole
t.test(data_ID$s_systole, data_ID$s_diastole, mu = 0, alt = "two.sided", conf = 0.95, paired = T)

# show means for systole and diastole
mean(data_ID$s_systole)
mean(data_ID$s_diastole)

## Plot sensitivity at systole and diastole in reference to an identity line

# save data as vectors
s_systole <- data_ID$s_systole
s_diastole <- data_ID$s_diastole

# generate square plotting region
par(pty="s")

# specify plotting space
plot(c(0,5),c(0,5),type="n",xlab="",ylab="", frame.plot = F,axes=F)

# specify axes
axis(2,c(0:4))
axis(1,c(0:4))

# plot the individual data
points(s_systole,s_diastole, pch = 1, col = rgb(0.23,0.48,0.50), cex = 1)

# prepare a matrix with sensitivity distribution (difference between sensitivity at systole and diastole)
diff <- as.matrix(data.frame(x=density(s_systole-s_diastole)$x,y=density(s_systole-s_diastole)$y))

# prepare rotation parameters
rotation_matrix <- matrix(c(cospi(1/4),sinpi(1/4),-sinpi(1/4),cospi(1/4)),ncol=2)

# rotate the matrix with sensitivity distribution 
diff_rotated <- diff %*% rotation_matrix

# plot the rotated sensitvity distribution
lines((diff_rotated)+3.2, col = rgb(0.23,0.48,0.50), lwd = 2)

# create the x axis
lines(c(1.8,4.5),c(4.5,1.8), col = "black", cex = 1)

# define the coordinates representing the values plotted the x axis
points_coordinates <- matrix(c(1,0,-1,0),nrow=2,byrow=T)

# rotate the coordinates
points_coordinates_rotated <- points_coordinates %*% rotation_matrix

# add the units to x axis
text(points_coordinates_rotated[1,1]+2.9,points_coordinates_rotated[1,2]+2.9,"1",srt=315,cex=.9)
text(points_coordinates_rotated[1,1]+3.1,points_coordinates_rotated[1,2]+3.1,"|",srt=315,cex=.53)
text(points_coordinates_rotated[2,1]+2.9,points_coordinates_rotated[2,2]+2.9,"-1",srt=315,cex=.9)
text(points_coordinates_rotated[2,1]+3.1,points_coordinates_rotated[2,2]+3.1,"|",srt=315,cex=.53)

# add the line endpoints 
text(points_coordinates_rotated[1,1]+3.8,points_coordinates_rotated[1,1]+1.1,"|",srt=315, cex= 0.9, col = "black")
text(points_coordinates_rotated[1,1]+1.09,points_coordinates_rotated[1,1]+3.8,"|",srt=315, cex= 0.9, col = "black")

# create an identity line using a customized linear model
x<-0:5
y<-0:5
new <- data.frame(x = seq(0, 4.5, 0.5))
lines(new$x, predict(lm(y~x), new),col= alpha("grey20", alpha = 0.5),lty= 2, lwd = 1.5)

# add axes labeles
title(xlab="Sensitivity at systole", mgp=c(2.3,0.5,0), ylab ="Sensitivity at diastole",adj= 0.32, cex = .8, cex.lab = 0.85)

# add legend
legend("topright", "sensitivity distribution" , lty=1, col=rgb(0.23,0.48,0.50), lwd = 1.5, bty='n', cex=.75)


```
<a name="S10"></a>
&nbsp;

**10. Test differences in response criterion between systole and diastole**

```{r}

## Response criterion (c) at systole and diastole

# test normality of criterion distribution at systole and diastole
shapiro.test(data_ID$c_systole)
shapiro.test(data_ID$c_diastole)

# employ paired t-test to determine whether the criterion differs between the systole and diastole
t.test(data_ID$c_systole, data_ID$c_diastole, mu = 0, alt = "two.sided", conf = 0.95, paired = T)

# show means for systole and diastole
mean(data_ID$c_systole)
mean(data_ID$c_diastole)

## Plot response criterion at systole and diastole in reference to an identity line

# save data as vectors (multiplying by 2 only for visualization purposes)
c_systole <- data_ID$c_systole*2
c_diastole <- data_ID$c_diastole*2

# generate square plotting region
par(pty="s")

# specify plotting space
plot(c(0,4.9),c(0,4.9),type="n",xlab="",ylab="", frame.plot = F,axes=F)

# specify customized axes
axis(1, at=0:4, labels=c("0","","1","","2"))
axis(2, at=0:4, labels=c("0","","1","","2"))

# plot the individual data
points(c_systole,c_diastole, pch = 1, col = rgb(0.23,0.48,0.50), cex = 1)

# prepare the matrix with response criterion distribution (difference between criterion at systole and diastole)
diff <- as.matrix(data.frame(x=density(c_systole-c_diastole)$x,y=density(c_systole-c_diastole)$y))

# prepare rotation parameters
rotation_matrix <- matrix(c(cospi(1/4),sinpi(1/4),-sinpi(1/4),cospi(1/4)),ncol=2)

# rotate the matrix with response criterion distribution
diff_rotated <- diff %*% rotation_matrix

# plot the rotated response criterion distribution
lines((diff_rotated)+3.3, col = rgb(0.23,0.48,0.50), lwd = 2)

# create the x axis
lines(c(1.9,4.6),c(4.6,1.9), col = "black", cex = 1)

# define the coordinates representing the values plotted the x axis
points_coordinates <- matrix(c(1,0,-1,0),nrow=2,byrow=T)

# rotate the coordinates
points_coordinates_rotated <- points_coordinates %*% rotation_matrix

# add the units to x axis
text(points_coordinates_rotated[1,1]+3.0,points_coordinates_rotated[1,2]+3.0,"0.5",srt=315,cex=.9)
text(points_coordinates_rotated[1,1]+3.2,points_coordinates_rotated[1,2]+3.2,"|",srt=315,cex=.53)
text(points_coordinates_rotated[2,1]+3.0,points_coordinates_rotated[2,2]+3.0,"-0.5",srt=315,cex=.9)
text(points_coordinates_rotated[2,1]+3.2,points_coordinates_rotated[2,2]+3.2,"|",srt=315,cex=.53)

# add the line endpoints
text(points_coordinates_rotated[1,1]+3.9,points_coordinates_rotated[1,1]+1.2,"|",srt=315, cex= 0.9, col = "black")
text(points_coordinates_rotated[1,1]+1.2,points_coordinates_rotated[1,1]+3.9,"|",srt=315, cex= 0.9, col = "black")
#plot(diff)

# create an identity line using a customized linear model
x<-0:5
y<-0:5
new <- data.frame(x = seq(0, 4.6, 0.2))
lines(new$x, predict(lm(y~x), new),col= alpha("grey20", alpha = 0.5),lty= 2, lwd = 1.5)

# add axes labeles
title(xlab="Criterion at systole", mgp=c(2.3,0.5,0), ylab ="Criterion at diastole",adj= 0.32, cex = .8, cex.lab = 0.85)

# add legend
legend(1, 5.3, "criterion distribution" , lty=1, col=rgb(0.23,0.48,0.50), lwd = 1.5, bty='n', cex=.75)


```
<a name="S11"></a>
&nbsp;



