---
title: <font size="5">**Interactions between cardiac activity and conscious somatosensory perception – supplement**</font>
author: <font size="4"> Pawe³ Motyka (pawel.motyka@psych.uw.edu.pl) </font>
date: <font size="2"> Semptember 25th 2018 </font>
output: html_document
chunk_output_type: console
---

&nbsp;
<font size="3">
**List of sections**:

1. Load and preprocess Interoceptive Accuracy data [S1](#S1)
2. Distribution of hits and misses for high and low interoceptive accuracy groups [S2](#S2)
3. Differences in hit rates between systole and diastole for high and low interoceptive accuracy groups [S3](#S3)

<a name="S1"></a>
&nbsp;

**1. Load and preprocess Interoceptive Accuracy data** </font>


```{r}

# Specify data directory
data_dir <- 'N:/CCSomato_data'
setwd(data_dir)

# load interoceptive accuracy data (Heartbeat Counting Task)
int_acc <- read.csv("CCSomato_HBT.csv", header = TRUE, sep = ";")

# load the preprocessed behavioral and ECG data
data <- read.table("CCSomato_ECG_data(output_from_section_4)", header = TRUE, sep = "\t", fill = TRUE, stringsAsFactors = FALSE)


# calculate averaged interoceptive accuracy from 5 intervals [1/5 sum of (1- (|recorded heartbeats – reported heartbeats|)/recorded heartbeats)]
int_acc$IA <- 0.2 * ((1 - ((abs(int_acc$interval1_measured - int_acc$interval1_counted))/int_acc$interval1_measured)) + (1 - ((abs(int_acc$interval2_measured - int_acc$interval2_counted))/int_acc$interval2_measured)) + (1 - ((abs(int_acc$interval3_measured - int_acc$interval3_counted))/int_acc$interval3_measured)) + (1 - ((abs(int_acc$interval4_measured - int_acc$interval4_counted))/int_acc$interval4_measured)) + (1 - ((abs(int_acc$interval5_measured - int_acc$interval5_counted))/int_acc$interval5_measured)) )

# perform exclusions
int_acc <- int_acc[int_acc$ID != 1,] # exclude participant 1 (one of the authors - piloting)
int_acc <- int_acc[int_acc$ID != 28,] # exclude participant 28 due to not properly reported heartbeats (all zereos)

# Reduce interoceptive accuracy data to ID and IA index
int_acc <- int_acc[, c(col=1, col=18)]

# Summary statistics
mean(int_acc$IA)
sd(int_acc$IA)
range(int_acc$IA)

# Create variable dividing participants into those with higher and lower interoceptive accuracy (by median)
median_IA <- median(int_acc$IA, na.rm = T)
median_IA

int_acc$IA_level[int_acc$IA <= median_IA] <- 'IA_lower' # lower interoceptive accuracy group
int_acc$IA_level[int_acc$IA > median_IA] <- 'IA_higher' # higher interoceptive accuracy group
int_acc$IA_level <- factor(int_acc$IA_level)

# merge data with interoceptive accuracy scores
data_Circ <- merge(data, int_acc, by = 'ID')

# Rayleigh Test yields significant result in case of participant nr 12 indicating a non-uniform distribution of stimulus onsets across the cardiac cycle (RR intervals)
# remove participant nr 12
data_Circ <- data_Circ[data_Circ$ID != 12,] 

```

<a name="S2"></a>
&nbsp;

<font size="3">
**2. Distribution of hits and misses for high and low interoceptive accuracy groups** </font>
&nbsp;

```{r}

require(circular, warn.conflicts = FALSE, quietly = T)
require(scales, warn.conflicts = FALSE, quietly=TRUE)

### Distribtuion of hits within RR intervals

# specify variables containig positions of hits within RR intervals
mean_degree_hit_IA_higher <- vector()
mean_degree_hit_IA_lower <- vector()

# extract the list of participants 
ID_list_IA_higher <- unique(data_Circ$ID[data_Circ$IA_level == "IA_higher"])
ID_list_IA_lower <- unique(data_Circ$ID[data_Circ$IA_level == "IA_lower"])


for (p in ID_list_IA_higher) { # LOOP PARTICIPANTS (p)
  
  # save the vector with hits positions in degrees as a circular object
  temp_hit <- circular(data_Circ$diff2peak_deg[data_Circ$ID==p & data_Circ$resp_class ==   "hit"], type="angles", units="degree", rotation="clock", zero=0)
  
  # calculate the mean phase where hits occurred for the participant
  mean_degree_hit_IA_higher[p] <- mean(temp_hit)
  
} # END: LOOP PARTICIPANTS (p)

for (p in ID_list_IA_lower) { # LOOP PARTICIPANTS (p)
  
  # save the vector with hits positions in degrees as a circular object
  temp_hit <- circular(data_Circ$diff2peak_deg[data_Circ$ID==p & data_Circ$resp_class ==   "hit"], type="angles", units="degree", rotation="clock", zero=0)
  
  # calculate the mean phase where hits occurred for the participant
  mean_degree_hit_IA_lower[p] <- mean(temp_hit)
  
} # END: LOOP PARTICIPANTS (p)


# save the vector with the mean hits for each particiapant as a circular object
Hits_secondlevel_IA_higher <- circular(mean_degree_hit_IA_higher[ID_list_IA_higher], type="angles", units="degree", rotation="clock", zero=pi/2)

Hits_secondlevel_IA_lower <- circular(mean_degree_hit_IA_lower[ID_list_IA_lower], type="angles", units="degree", rotation="clock", zero=pi/2)

# plot the distribution of mean hits within RR interval across participants - with 0 indicating the R peak previous to the stimulus onset
plot(Hits_secondlevel_IA_higher, stack= TRUE , bins = 720, col = alpha("gray10", 0.8), cex = 0.88, lwd = 2, main = "Hits")
points(Hits_secondlevel_IA_lower, col = "gray25", cex = 0.77, lwd = 2)

# add arrows representing mean hits for each participant
arrows.circular(Hits_secondlevel_IA_higher, col =  rgb(0.55, 0.23, 0.23, 0.25), lwd = 3, length = 0.001)
arrows.circular(Hits_secondlevel_IA_lower, col = rgb(0.247, 0.435, 0.545, 0.25), lwd = 3, length = 0.001)

# add an arrow representing grand mean for hits - with its length signifying resultant length of the grand mean
arrows.circular(mean(Hits_secondlevel_IA_higher), y=rho.circular(Hits_secondlevel_IA_higher), lwd = 3, col = rgb(0.45, 0.13, 0.13, 0.75), length = 0.08)
arrows.circular(mean(Hits_secondlevel_IA_lower), y=rho.circular(Hits_secondlevel_IA_lower), lwd = 3, col = rgb(0.147, 0.285, 0.395, 0.75), length = 0.08)

# add line signifying distribution of hits (bandwidth = 20)
circ.dens = density(Hits_secondlevel_IA_higher, bw=20)
lines(circ.dens, col= rgb(0.55, 0.23, 0.23), lwd = 1.5, xpd=TRUE)

circ.dens = density(Hits_secondlevel_IA_lower, bw=20)
lines(circ.dens, col= rgb(0.247, 0.435, 0.545), lwd = 1.5, xpd=TRUE)

# show the angle of grand mean for hits
mean(Hits_secondlevel_IA_higher)
mean(Hits_secondlevel_IA_lower)

# Test for a Common Distribution - two independent samples (p < 0.05 - interpreted as an evidence that directions have been drawn from different distributions)
watson.two.test(Hits_secondlevel_IA_higher, Hits_secondlevel_IA_lower, alpha = 0.05)

# test uniformity of hits distribution within RR intervals
rayleigh.test(Hits_secondlevel_IA_higher)
rayleigh.test(Hits_secondlevel_IA_lower)


### Distribtuion of misses within RR interval

# specify variables containig positions of misses within RR intervals
mean_degree_miss_IA_higher <- vector()
mean_degree_miss_IA_lower <- vector()

for (p in ID_list_IA_higher) { # LOOP PARTICIPANTS (p)
  
  # save the vector with misses positions in degrees as a circular object
  temp_miss <- circular(data_Circ$diff2peak_deg[data_Circ$ID==p & data_Circ$resp_class == "miss"], type="angles", units="degree", rotation="clock", zero=0)
  
  # calculate the mean phase where misses occurred for the participant
  mean_degree_miss_IA_higher[p] <- mean(temp_miss)
  
} # END: LOOP PARTICIPANTS (p)

for (p in ID_list_IA_lower) { # LOOP PARTICIPANTS (p)
  
  # save the vector with misses positions in degrees as a circular object
  temp_miss <- circular(data_Circ$diff2peak_deg[data_Circ$ID==p & data_Circ$resp_class == "miss"], type="angles", units="degree", rotation="clock", zero=0)
  
  # calculate the mean phase where misses occurred for the participant
  mean_degree_miss_IA_lower[p] <- mean(temp_miss)
  
} # END: LOOP PARTICIPANTS (p)

# save the vector with the mean misses for each particiapant as a circular object
misses_secondlevel_IA_higher <- circular(mean_degree_miss_IA_higher[ID_list_IA_higher], type="angles", units="degree", rotation="clock", zero=pi/2)

misses_secondlevel_IA_lower <- circular(mean_degree_miss_IA_lower[ID_list_IA_lower], type="angles", units="degree", rotation="clock", zero=pi/2)

# plot the distribution of mean misses within RR interval across participants - with 0 indicating the R peak previous to the stimulus onset
plot(misses_secondlevel_IA_higher, stack=TRUE, bins = 720, col = alpha("gray10", 0.8), cex = 0.88, lwd = 2, main = "Misses")
points(misses_secondlevel_IA_lower, col = "gray25", cex = 0.77, lwd = 2)

# add arrows representing mean Misses for each participant
arrows.circular(misses_secondlevel_IA_higher, col =  rgb(0.55, 0.23, 0.23, 0.25), lwd = 3, length = 0.001)
arrows.circular(misses_secondlevel_IA_lower, col = rgb(0.247, 0.435, 0.545, 0.25), lwd = 3, length = 0.001)

# add an arrow representing grand mean for Misses - with its length signifying resultant length of the grand mean
arrows.circular(mean(misses_secondlevel_IA_higher), y=rho.circular(misses_secondlevel_IA_higher), lwd = 3, col = rgb(0.45, 0.13, 0.13, 0.76), length = 0.08)
arrows.circular(mean(misses_secondlevel_IA_lower), y=rho.circular(misses_secondlevel_IA_lower), lwd = 3, col = rgb(0.147, 0.285, 0.395, 0.76), length = 0.08)

# add line signifying distribution of Misses (bandwidth = 20)
circ.dens = density(misses_secondlevel_IA_higher, bw=20)
lines(circ.dens, col=rgb(0.55, 0.23, 0.23), lwd = 1.5, xpd=TRUE)

circ.dens = density(misses_secondlevel_IA_lower, bw=20)
lines(circ.dens, col=rgb(0.247, 0.435, 0.545), lwd = 1.5, xpd=TRUE)

# show the angle of grand mean for Misses
mean(misses_secondlevel_IA_higher)
mean(misses_secondlevel_IA_lower)

# Test for a Common Distribution - two independent samples (p < 0.05 - interpreted as an evidence that directions have been drawn from different distributions)
watson.two.test(misses_secondlevel_IA_higher, misses_secondlevel_IA_lower, alpha = 0.05)

# test uniformity of misses distribution within RR intervals
rayleigh.test(misses_secondlevel_IA_higher)
rayleigh.test(misses_secondlevel_IA_lower)


### Test correlation between mean resultant lengths (i.e., the level of hits or misses concentration) and interoceptive accuracy

## Hits

Hits <- data.frame(concentration = numeric(0),
                             IA = numeric(0))

for (p in unique(data_Circ$ID)) { # LOOP PARTICIPANTS (p)
  
  # save the vector with hits positions in degrees as a circular object
  temp_hit <- circular(data_Circ$diff2peak_deg[data_Circ$ID==p & data_Circ$resp_class == "hit"], type="angles", units="degree", rotation="clock",    zero=0)
  
  # save the results of Rayleigh test
  hits_distribution_individual <- rayleigh.test(temp_hit) 
  
  # save the value of mean resultant length 
  concentration <- hits_distribution_individual$statistic
  
  # save IA score
  IA <- unique(data_Circ$IA[data_Circ$ID == p])
  
  # create data frame
   Hits[nrow(Hits)+1,] <- c(concentration, IA)
  
}

# test normality of the variables
shapiro.test(Hits$IA)
shapiro.test(Hits$concentration)

# test correlation between hits concetration and interoceptive accuracy 
cor.test(Hits$IA, Hits$concentration, method ="pearson")

## Misses

Misses <- data.frame(concentration = numeric(0),
                             IA = numeric(0))

for (p in unique(data_Circ$ID)) { # LOOP PARTICIPANTS (p)
  
   # save the vector with Misses positions in degrees as a circular object
  temp_hit <- circular(data_Circ$diff2peak_deg[data_Circ$ID==p & data_Circ$resp_class == "miss"], type="angles", units="degree", rotation="clock", zero=0)
  
  # save the results of Rayleigh test
  Misses_distribution_individual <- rayleigh.test(temp_hit) 
  
  # save the value of mean resultant length
  concentration <- Misses_distribution_individual$statistic
  
  #  save IA score
  IA <- unique(data_Circ$IA[data_Circ$ID == p])
  
  # create data frame
   Misses[nrow(Misses)+1,] <- c(concentration, IA)
  
}

# test normality of the variables
shapiro.test(Misses$IA)
shapiro.test(Misses$concentration)

# test correlation between Misses concetration and interoceptive accuracy 
cor.test(Misses$IA, Misses$concentration, method ="pearson")



```

<a name="S3"></a>

&nbsp;

<font size="3">
**3. Differences in hit rates between systole and diastole for high and low interoceptive accuracy groups** </font> &nbsp;
```{r, fig.height= 5, fig.width= 5}

# load preprocessed data
data <- read.table("CCSomato_ECG_data(output_from_section_7)", header = TRUE, sep = "\t", fill = TRUE, stringsAsFactors = FALSE)

require(ggplot2, warn.conflicts = FALSE, quietly=TRUE)
require(effsize, warn.conflicts = FALSE, quietly=TRUE)
require(scales, warn.conflicts = FALSE, quietly=TRUE)
require(car, warn.conflicts = FALSE, quietly=TRUE)
require(ez, warn.conflicts = FALSE, quietly=TRUE)
require(tidyr, warn.conflicts = FALSE, quietly=TRUE)

data_ID <- data.frame(ID = integer(0),
                 hit_rate = numeric(0),
                 hit_rate_s = numeric(0), # hit rate at systole
                 hit_rate_d = numeric(0)) # hit rate at diastole


for ( p in unique(data$ID)) { # LOOP PARTICIPANTS (p)
    
    # calculate hit rate
    hit_rate <- (length(data$resp_class[data$resp_class =="hit" & data$ID == p]))/(length(data$trial[data$ID == p & data$stim_type == 1]))
    
   
    # calculate hit rate at systole
    hit_rate_s <- (length(data$resp_class[data$resp_class =="hit" & data$ID == p & data$systole == T]))/(length(data$trial[data$ID == p & data$stim_type == 1 & data$systole == T]))
    
    # calculate hit rate at diastole
    hit_rate_d <- (length(data$resp_class[data$resp_class =="hit" & data$ID == p & data$systole == F]))/(length(data$trial[data$ID == p & data$stim_type == 1 & data$systole == F]))

    # add calculated varialbes to the data frame
    data_ID[nrow(data_ID)+1,] <- c(p, hit_rate, hit_rate_s,hit_rate_d)
    
} # END: LOOP PARTICIPANTS (p)

data_ID$hit_rate <- data_ID$hit_rate * 100
data_ID$hit_rate_s <- data_ID$hit_rate_s * 100
data_ID$hit_rate_d <- data_ID$hit_rate_d * 100

# test normality of hit rate distribution at systole and diastole
shapiro.test(data_ID$hit_rate_s)
shapiro.test(data_ID$hit_rate_d)

# merge data frames
data_ID <- merge(int_acc, data_ID, by = 'ID')


### Run Two-way mixed ANOVA

data_ID <- data_ID[, c(col=c(1,3,5,6))]

dat <- gather(data_ID, phase, hit_rate, hit_rate_s:hit_rate_d, factor_key = T)
dat$ID <- as.factor(dat$ID)

#ezAnova method
anova_ez = ezANOVA(data=dat, wid=ID, dv = hit_rate, within = phase, between=IA_level, detailed = TRUE)
print(anova_ez)

### Plot

# save data as vectors
s_hit_rate_high <- data_ID$hit_rate_s[data_ID$IA_level == "IA_higher"] 
d_hit_rate_high <- data_ID$hit_rate_d[data_ID$IA_level == "IA_higher"] 
s_hit_rate_low <- data_ID$hit_rate_s[data_ID$IA_level == "IA_lower"] 
d_hit_rate_low <- data_ID$hit_rate_d[data_ID$IA_level == "IA_lower"] 

par(pty="s")

# specify plotting space
plot(c(0,109),c(0,109),type="n",xlab ="", ylab="", cex.lab = 0.8, cex.axis = 0.8, frame.plot = F)

title(ylab="Hit rate at systole (%)", line = 2.5, cex.lab= 0.8)
title(xlab="Hit rate at diastole (%)", line = 2.5, cex.lab= 0.8)

# create an identity line using a customized linear model
x<-0:100
y<-0:100
new <- data.frame(x = seq(0, 100, 0.5))
lines(new$x, predict(lm(y~x), new),col= alpha("black", alpha = 0.8),lty= 2, lwd = 1.3)

# plot the individual data
points(d_hit_rate_high, s_hit_rate_high, pch = 1, col = rgb(0.55, 0.23, 0.23), cex = 1) # higher IA group
points(d_hit_rate_low, s_hit_rate_low, pch = 1, col = rgb(0.247, 0.435, 0.545), cex = 1) # lower IA group


# prepare the matrix with sensitivity distribution (difference between sensitivity at systole and diastole)
diff_high <- as.matrix(data.frame(x=density(d_hit_rate_high-s_hit_rate_high)$x,y=density(d_hit_rate_high-s_hit_rate_high)$y))
diff_low <- as.matrix(data.frame(x=density(d_hit_rate_low-s_hit_rate_low)$x,y=density(d_hit_rate_low-s_hit_rate_low)$y))

# rescale y axis to match size of the plot
f <- scales::rescale(diff_high[,2], to = c(0, 20))
diff_high <- as.matrix(data.frame(x=density(d_hit_rate_high-s_hit_rate_high)$x,y=f))

f2 <- scales::rescale(diff_low[,2], to = c(0, 20))
diff_low <- as.matrix(data.frame(x=density(d_hit_rate_low-s_hit_rate_low)$x,y=f2))

# prepare rotation parameters
rotation_matrix <- matrix(c(cospi(1/4),sinpi(1/4),-sinpi(1/4),cospi(1/4)),ncol=2)

# rotate the matrix with sensitivity distribution 
diff_high_rotated <- diff_high %*% rotation_matrix
diff_low_rotated <- diff_low %*% rotation_matrix

# plot hit rate distribution
lines((diff_high_rotated)+90, col = rgb(0.55, 0.23, 0.23), lwd = 1.7)
lines((diff_low_rotated)+90, col = rgb(0.247, 0.435, 0.545), lwd = 1.7)

# create the x axis
lines(c(74,103.4),c(103.4,74), col = "black", cex = 1)

# define the coordinates representing the values plotted on the x axis
points_coordinates <- matrix(c(1,0,-1,0),nrow=2,byrow=T)

# rotate the coordinates
points_coordinates_rotated <- points_coordinates %*% rotation_matrix

# add units to x axis
text(points_coordinates_rotated[1,1]+92,points_coordinates_rotated[1,2]+77,"+10%",srt=315,cex=.7)
text(points_coordinates_rotated[1,1]+103,points_coordinates_rotated[1,2]+74.7,"|",srt=315,cex=.8)
text(points_coordinates_rotated[2,1]+77,points_coordinates_rotated[2,2]+92,"-10%",srt=315,cex=.7)
text(points_coordinates_rotated[2,1]+74.7,points_coordinates_rotated[2,2]+103,"|",srt=315,cex=.8)

# add endpoints to x axis
text(points_coordinates_rotated[1,1]+94.7,points_coordinates_rotated[1,1]+80,"|",srt=315, cex= 0.35, col = "black")
text(points_coordinates_rotated[1,1]+80,points_coordinates_rotated[1,1]+94.7,"|",srt=315, cex= 0.35, col = "black")


```





